\documentclass[11pt]{article}
\usepackage[utf8]{inputenc}
\usepackage[a4paper, margin=1in]{geometry}
\usepackage{amsmath}    % advanced math environments
\usepackage{amssymb}    % mathematical symbols
\usepackage{amsfonts}
\usepackage{graphicx}   % include images
\usepackage{booktabs}   % professional tables
\usepackage{hyperref}   % clickable references and links
\usepackage[backend=biber, style=apa]{biblatex}
\addbibresource{reference.bib}

\title{Impact of early-phase design strategies on overall drug development success rate}
\author{Yongxi Long \and Rob Kessels \and Sama Heydari \and Elias Sjoerd \and Peter van de Ven}
\date{\today}

\begin{document}
\maketitle
\begin{abstract}
A considerable amount of effort has been dedicated to improve the design of early-phase dose-finding trials in oncology. However, few of them had applications to real trials and the conventional yet inefficient 3+3 design remains a common choice. Most comparative studies of early-phase trial designs focus on Phase I operating characteristics, leaving open the question of whether differences in dose selection meaningfully translate into downstream impact on drug development success.

We propose a simulation framework to evaluate how representative dose-finding methods (3+3, BOIN, and CRM) impact the probability of significant results in Phase III trials, representing overall drug development success. We consider both chemotherapy-like profiles, where the classical monotonicity assumption about toxicity and efficacy holds, and immunotherapy-like drug profiles where the efficacy typically plateaus before reaching the maximum tolerable dose (MTD). Different toxicity control rules are implemented in Phase II and III to reflect varying extent of safety monitoring in practice. We further compare univariate designs to their bivariate extensions to investigate the potential benefits of incorporating early efficacy data. Finally, we redesign the development process of a targeted therapy, Imatinib in patients with advanced gastrointestinal stromal tumour (GIST), to illustrate the proposed simulation pipeline. 

For chemotherapy-like drug profiles, BOIN and CRM substantially outperforms 3+3 across a range of late-phase toxicity control settings. In contrast, for immunotherapy-like drug profiles, the advantage of more efficient designs over 3+3 is compromised, particularly under stricter toxicity control. Bivariate designs, such as EffTox and BOIN12, yield more robust overall success rate. BOIN12 is less sensitive to utility tuning parameters compared to EffTox.

The downstream performance of dose-finding methods heavily depends on the underlying drug profiles. Designs that perform well in identifying the MTD leads to a poor late-phase success rate when the dose-efficacy relationship has an early plateau in immunotherapy-like profiles. Utilizing efficacy information in Phase I does not necessarily guarantee higher overall success rates due to small sample sizes and insufficient follow-up in early phases. These findings highlight the importance to understand the potential shape of the candidate drug profile from pre-clinical and pilot studies when designing early-phase trials.
\end{abstract}
\section{Introduction}
The past decades have witnessed the approval of many effective therapies based on clinical trial evidence, especially in the field of oncology. Early-phase dose-findings trials play a critical role in characterizing the safety and dosing of promising therapeutic candidates, but also come with great uncertainty because of small sample sizes. Despite the breakthrough in pharmaceutical research, the overall success rate of drug development remains low, with approximately $90\%$ of drug candidates eventually failing after advacing to the Phase I stage. One of the major reasons for the high attrition rate is the lack of clinical efficacy and unmanageable toxicity, which can often be attributed to suboptimal dose selections in early-phase trials \parencite{drug_failure}. 

Traditionally, dose selection has been guided by the toxicity-efficacy relationship observed in chemotherapies, where the monotonicity assumption typically holds. Trial designs escalate doses until reaching the maximum tolerable dose (MTD). However, the landscape of oncology drug development has been transformed by the emergence of novel immunotherapies. For these drugs, efficacy often plateaus before reaching the MTD, necessitating consideration of both toxicity and efficacy to identify an optimal biological dose (OBD) \parencite{MTD_OBD, OBD_clinical_efficacy}. Further complexities arise in settings with patient heterogeneity \parencite{wages2015phase,guo2022bayesian,guo2022bipse}, delayed outcomes \parencite{TITE-CRM,TITE-BOIN12}, and unclear dose ordering in combination therapies \parencite{thall2003dose,BOINcomb}, etc. Consequently, substantial effort has been devoted to optimize the dose selection in early-phase trials (Kessels at. al review).

Despite extensive methodological development, the adoption of novel statistical designs in clinical practice has been slow: the conventional yet inefficient 3+3 design remains a common choice in approximately 74\% of oncology Phase I trials \parencite{review_phase1}. The popularity of the 3+3 design comes primarily from its simplicity. As a rule-based design, it follows straightforward escalation and de-escalation rules and requires minimal prior knowledge \parencite{rule_based_design}. 

However, rule-based designs only use data from the immediately previous cohort to guide dose decisions, discarding potentially valuable information \parencite{clertant2022early}. To address this limitation, model-based designs such as the Continuous Reassessment Method (CRM) were developed to use all accumulated data and explicitly estimate the underlying dose profile \parencite{CRM,EWOC,review_model_based_design}. While being more efficient, these approaches are also more complex. Model-assisted designs, such as the Bayesian Optimal Interval (BOIN) design \parencite{BOIN}, strike a balance by using all data from the current dose level while maintaining implementation simplicity comparable to rule-based designs \parencite{review_model_assisted_design}. Despite their advantages, model-assisted and model-based designs still encounter practical complexities of implementation: they often require trial-specific boundaries and/or initial guess of the drug profile, and more involvement of statistical expertise \parencite{clertant2022early}. Additional barriers include the lack of practical implementation tools, limited real-world experience, and underdeveloped guidance for protocol development \parencite{petroni2017implementation}.

However, beyond implementation challenges, there remains also a gap in how these designs are evaluated in the literature. Most comparative studies of early-phase trial designs focus exclusively on Phase I operating characteristics, such as the probability of correctly identifying the MTD and the probability of overdosing \parencite{horton2017performance,iasonos2008comprehensive,van2016performance,zhou2018comparative}. While these metrics are information, they do not capture the ultimate goal of drug development: bringing safe and efficacious treatments to patients. The extent to which performance differences across dose-finding designs translate into meaningful downstream outcomes, particularly the probability of success in late-phase trials, remains largely unexamined.

Some studies have attempted to evaluate the downstream consequences of dose selection. \textcite{conaway2019impact} reported that CRM cnsistently yielded a higher probability of success in Phase III trials compared to 3+3 and BOIN. However, the drug profiles considered exhibited relatively low toxicity, which may not capture the aggressive escalation behaviour of CRM, a design known to be prone to overdosing \parencite{EWOC}. It is therefore unclear whether the advatange of CRM is generalizable beyond this restricted class of profiles. In addition, excess toxicity was assessed using a binomial test at the end of the Phase III trial, which may not reflect the real-world practice of continous safety monitoring and is likely to be underpowered. \textcite{ruppert2018overall} evaluated the overall drug development success rate across a range of dose-toxicity and dose-efficacy relationships. They found heterogenous performance of dosing-finding methods depending on the shape of the underlying drug profiles. But no toxicity control rules were implemented in the late-phase trials, potentially offfering unrealistic advantage to more aggressive dose-finding methods. Morever, both studies focused primarily on univariate designs that rely solely on toxicity data to guide dose selection. This focus may be increasingly less relevant for current practice in oncology, given the growing emphasis towards OBD selection in the development of non-cytotoxix agents \parencite{araujo2023oncology}. 

In this paper, we propose a comprehensive simulation framework to evaluate the impact of representative dose-finding methods on the success rate of Phase III trials. We assess the robustness of each method across various drug profiles featured by both chemotherapies, where the dose-efficacy relationship parallels the dose-toxicity relationship, and immunotherapies, where the efficacy plateaus before the MTD. We further consider different toxicity control rules in late-phase trials to reflect varying extent of safety monitoring in practice. We addtionally compare univariate designs to their bivariate extensions to quantify the potential benefit and limitations of incorporating early efficacy data in dose escalation. Finally, we redesign the development process of a targeted therapy, Imatinib in patients with advanced gastrointestinal stromal tumour (GIST), to illustrate the proposed simulation pipeline.

\newpage
\addcontentsline{toc}{section}{Reference}
\section*{Reference}
\printbibliography [heading=none]
\end{document}